Improving Code Completion in Pharo\\Using N-gram Language Models

### 2020-05-15 ACM/abstract.tex
about halfway rewritten

\documentclass[format=sigconf]{acmart}

\usepackage{xcolor} %%% for comments
\usepackage{ifthen} %%% for comments
\usepackage{ulem} %%% for striked out text

% comments \nb{label}{color}{text}
\newboolean{showcomments}
\setboolean{showcomments}{true}
\ifthenelse{\boolean{showcomments}}
	{\newcommand{\nb}[3]{
		{\colorbox{#2}{\bfseries\sffamily\scriptsize\textcolor{white}{#1}}}
		{\textcolor{#2}{\sf\small$\blacktriangleright$\textit{#3}$\blacktriangleleft$}}}
	 \newcommand{\version}{\emph{\scriptsize$-$Id$-$}}}
	{\newcommand{\nb}[3]{}
	 \newcommand{\version}{}}
\newcommand{\md}[1]{\nb{Marcus}{red}{#1}}
\newcommand{\oz}[1]{\nb{Oleks}{olive}{#1}}
\newcommand{\mr}[1]{\nb{Myroslava}{teal}{#1}}

\title{Improving Code Completion in Pharo\\Using N-gram Language Models}
%\title{Usage of N-gram models to improve autocompletion in Pharo}
\author{Myroslava Romaniuk}
\affiliation{\institution{Ukrainian Catholic University}}
\email{romaniuk@ucu.edu.ua}

\begin{document}

\begin{abstract}
In this paper we present applying statistical language models to improve code completion
in Pharo \footnote{https://pharo.org/}. In particular, the goal is to use n-gram models for
sorting the completion candidates and in such a way increase the relevancy of untyped suggestions.

\section{Introduction}
Code completion is one of the essential features of any IDE that greatly improves the
developer experience and productivity. Thus, it is important to find a way to make it as effective as
possible. Specifically, in the case of Pharo, a dynamically-typed Smalltalk-based programming language
and IDE, in the last year and a half a lot of work has been carried out to improve code
completion. However, there are more potential improvements from which developers can only
benefit, such as improving the relevance of suggestions and hence reducing time and
effort the developer spends to go through the list of completions.

\section{Context}
The current implementation of code completion in Pharo is based on abstract syntax tree
(AST) information of source code, where we are able to determine the structure of the code
and infer the type information where possible. Once we have a list of semantically suitable
completion candidates, they are sorted alphabetically. However, this approach ignores the
potential of suggesting completion results based on code history, which is something that
has been successfully adopted by other IDEs. In this work I intend to study how code
completion in Pharo can be improved using machine learning (ML) and statistical language
models. In particular, the idea is to use the N-gram model, as it has been documented to be
used for such a task \cite{Hind12a}, and, if successfully adapted to Pharo, might help
enhance the quality of code completion.

There are many approaches to handling the process of completing code. The classic strategies
use lexical \footnote{https://code.visualstudio.com/docs/editor/intellisense} (completing
based on the prefix the user is typing) or semantic (use code structure analysis information
to tailor completions) models. In the latest years statistical and ML models have also gained
popularity. For the most part when using ML, source code analysis is also often disregarded,
and the results of simply training the model on the code base and inferring all the possible
code dependencies are used instead.

\section{Related Works}
My work was inspired by the paper "On The Naturalness of Software" \cite{Hind12a}. There, as well
as in \cite{Tu14a,Rayc14a,Hell17a,Li17a}, the structural information is not taken into consideration
and the problem of improving code completion is reduced to a natural language processing problem,
where predicting the next token is treated as predicting the next word in a sentence.

@@

\section{Our Approach}
What we want to do in Pharo is slightly different. Unlike the implementations of \cite{Tu14a,Hind12a,Rayc14a,Hell17a,Li17a},
where code completion is based solely on the results of statistical or machine learning models, the code
completion in Pharo works by making use of AST information and giving contextually relevant results.
However, as Pharo is a dynamically-typed language, type information is not always available, and the
suggested results, after being sorted alphabetically, require more effort to go through and manually
narrow down. This is where we think a sorting strategy based on an n-gram model will work better. In this
case tokens will be suggested according to their probability of appearing in the already typed sequence,
which will allow us to propose more accurate suggestions even without knowing the exact type of the token.

@@

@@

\end{abstract}

\maketitle

\bibliographystyle{ACM-Reference-Format}
\bibliography{rmod,others}

\end{document}

>>>>

\documentclass[sigconf,screen]{acmart}

\title{Improving Code Completion in Pharo\\Using N-gram Language Models}
\author{Myroslava Romaniuk}
\affiliation{\institution{Ukrainian Catholic University}}
\email{romaniuk@ucu.edu.ua}

%%% The following is specific to <Programming'20> Companion-SRC and the paper
%%% 'Improving Code Completion in Pharo using N-gram Language Models'
%%% by Myroslava Romaniuk.
%%%
\setcopyright{acmcopyright}
\acmPrice{15.00}
\acmDOI{10.1145/3397537.3398483}
\acmYear{2020}
\copyrightyear{2020}
\acmSubmissionID{prog20src-p8-p}
\acmISBN{978-1-4503-7507-8/20/03}
\acmConference[<Programming'20> Companion]{Companion Proceedings of the 4th International Conference on the Art, Science, and Engineering of Programming}{March 23--26, 2020}{Porto, Portugal}
\acmBooktitle{Companion Proceedings of the 4th International Conference on the Art, Science, and Engineering of Programming (<Programming'20> Companion), March 23--26, 2020, Porto, Portugal}

\begin{document}
\begin{abstract}
In this paper I present applying statistical language models to improve code completion in Pharo\footnote{\url{https://pharo.org/}}. In particular, the goal is to use n-gram models for sorting the completion candidates and in such a way increase the relevancy of the suggested completions.

\section{Introduction}
Code completion is one of the most essential features in any IDE; it is one of the first things a developer notices and it is something that can "make or break" the flow of productivity when coding. The accuracy and the speed with which the completions are suggested are paramount in helping make the completion as effective as possible. That is why researchers and software engineers are constantly trying to find ways to enhance completion engines in IDEs.

When it comes to code completion in the Pharo IDE\footnote{Pharo is an object-oriented dynamically typed programming language inspired by Smalltalk, and the Pharo IDE is an interactive IDE intended for developing in Pharo.}, the situation is no different: over the past couple of years a lot of work has already been done to improve the completion engine there. However, I believe that it can only benefit from additionally improving the sorting strategies by using machine learning techniques.

\section{Context}
The current implementation of code completion in the Pharo IDE is based on abstract syntax tree (AST) information of source code, where we are able to determine the structure of the code and infer the type information where possible. We can then give contextually suitable completions, such as suggesting a class name for a global node, method names for a method node, and so on. However, once we get the list of suggestions, there is no efficient way to sort them, and so the desired completions can appear at the bottom of the list. This requires the developer to scroll down or type in more symbols and can be very unproductive.

In this work I intend to study how code completion in Pharo can be improved using statistical language models. In particular, the idea is to use the n-gram model, as it has been documented to be used for such a task \cite{Hind12a}, and, if successfully adapted to Pharo, I believe it can help enhance the quality of code completion.

\section{Related Works}
There are many approaches to handling the process of completing code. The classic strategies used to rely on language-specific pattern matching, i.e. completing based on the prefix the developer has typed in. According to Bruch et al. \cite{Bruc09a}, code completion engines up until 10 years ago would also mainly rely on type information, with no contextual analysis. In the paper, the authors countered this approach by implementing intelligent code completion systems, which learned from source code examples and gave much more contextually relevant results. In the latest years statistical and machine learning models have also gained popularity. In machine learning approaches, source code is often treated as regular text, and the contextual analysis is disregarded. IntelliSense\footnote{\url{https://code.visualstudio.com/docs/editor/intellisense}} is an example of a code completion tool that uses semantic analysis, whereas TabNine\footnote{\url{https://www.tabnine.com/}} is an example of a tool using deep learning.

This work has been inspired by the paper "On The Naturalness of Software" \cite{Hind12a}. There, as well as in \cite{Tu14a,Rayc14a,Hell17a,Li17a}, the structural information is not taken into consideration and the problem of improving code completion is reduced to a natural language processing problem, where predicting the next token is treated as predicting the next word in a sentence.

@@

\section{Proposed Approach}
\subsection{Implementation}
What we want to do in Pharo is slightly different. Unlike the implementations of \cite{Tu14a,Hind12a,Rayc14a,Hell17a,Li17a}, where code completion is based solely on the results of statistical or machine learning models, the code completion in Pharo works by making use of AST information and giving contextually relevant results. However, as Pharo is a dynamically-typed language, type information is not always available, and the suggested results, after being sorted alphabetically, require more effort to go through and manually narrow down. This is where we think a sorting strategy based on an n-gram model will work better. In this case tokens will be suggested according to their probability of appearing in the already typed sequence, which will allow us to propose more accurate suggestions even without knowing the exact type of the token.

@@

\subsection{Evaluation}

@@

\end{abstract}
\maketitle
\bibliographystyle{ACM-Reference-Format}
\bibliography{rmod,others}
\end{document}
